Cultural Transmission Can Explain the Evolution of Cooperation

### 2020-07-28 ms/main.tex
minor fix
continued revision of result 2

%%% Figure: model illustration
\begin{figure}[b]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}
%%%

@@

which gives the following result.

\begin{result}[Oblique and horizontal transmission of cooperation]
If there is a horizontal transmission bias in favor of cooperation, namely
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).
\end{result}

@@

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
p(1-p)\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.

Therefore, for cooperation to take over the population ($\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$, that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_global_condition}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1 .
\end{equation}

If $\alpha<1$, we want to determine a condition for $\tilde{p}'>\tilde{p}$ in \autoref{eq:nextgen_parents_vertical_only}. 
We divide  by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$ to get
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Rearranging, we get
\begin{equation} 
\begin{aligned} 
  1 - \tilde{p} < 
  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
We divide by $(1-\tilde{p})$ and rearrange so that free terms are on the left and terms with $\tilde{p}$ are on the right. We get
\begin{equation} 
\begin{aligned} 
  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
   + (1-c) (1-\alpha) T_B
   + (1+b) (1-\alpha) T_A 
   - (1-\alpha) T_A].
\end{aligned}
\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

Therefore, in this case there can be a third, polymorphic equilibrium, if $0<\tilde{p}^*<1$, where
\begin{equation} \label{eq:vert_hori_equilibrium}
  \tilde{p}^* = 
  \frac{c(1-T_B) - b \alpha T_A - (T_A-T_B)}{b (1-\alpha) (T_A-T_B) }.
\end{equation}

Note that all parameters are positive.
So, applying \autoref{eq:vert_hori_global_condition}, for $\tilde{p}'>\tilde{p}$,  we require that either 
\begin{equation} \label{eq:vert_hori_global_condition1}
T_A>T_B, \;\text{and}\; \tilde{p}>\tilde{p}^*,
\end{equation} or 
\begin{equation} \label{eq:vert_hori_global_condition2}
T_A<T_B, \;\text{and}\; \tilde{p}<\tilde{p}^* .
\end{equation}

@@

\begin{result}[Vertical and horizontal transmission of cooperation]
When $0 \le \alpha<1$ and $v=1$, fixation, extinction, and stable coexistence of both cooperation and defection are possible.
\end{result}

Determining the outcome of the dynamics depends on the sign of $T_A-T_B$ and on the legitimacy of the third equilibrium, that is, if $0<\tilde{p}^*<1$.
\\

%This condition can also be formulated as a condition on the cost of cooperation $c$: cooperation will take over the population if
%\begin{equation} \label{eq:unequal_transmission}
%c < \frac{
%	b \big[\alpha T_A + (1-\alpha)\tilde{p}(T_A - T_B)\big] + (T_A - T_B)
%	}{
%	1-T_B
%	}.
%\end{equation}
%
%Importantly, the condition for cooperation to increase from rarity is (set $\tilde{p}=0$ in \autoref{eq:unequal_transmission})
%\begin{equation} \label{eq:unequal_transmission_from_rarity}
%c < \frac{
%	b \alpha T_A + (T_A - T_B)
%	}{
%	1-T_B
%	} .
%\end{equation} 
%
%In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
%Without horizontal transmission bias, we get the following corollary that does have the form of Hamilton's rule.

@@

\begin{corollary}[No assortment of transmission and cooperation]
If $\alpha=0$, then the third equilibrium is
\begin{equation} \label{eq:vert_hori_alpha0_equilibrium}
\tilde{p}^*(\alpha=0) = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B) }.
\end{equation} 
If there is a horizontal transmission bias for cooperation, $T_A>T_B$, then cooperation will fix in the population if the cost is not too high $c < (1+b) \frac{T_A-T_B}{1-T_B}$, and the initial frequency is high enough, $\tilde{p}(0) > \tilde{p}^*$.
If there is a horizontal transmission bias for defection, $T_A<T_B$, defection will take over the population.
\end{corollary}

This can be shown by setting $\alpha=0$ in \autoref{eq:vert_hori_equilibrium}. 
Then, when $T_A>T_B$ we find a condition for $\tilde{p}^*<1$ and $\tilde{p}>\tilde{p}^*$ (\autoref{eq:vert_hori_global_condition1}). When $T_A<T_B$, we have $\tilde{p}^*<0$ and because of \autoref{eq:vert_hori_global_condition2}, any initial frequency leads to fixation of defection, $\tilde{p}=0$.\\

>>>>

%%% Figure: model illustration
\begin{figure}[thb]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}
%%%

@@

which gives the following result.\\

\begin{result}[Oblique and horizontal transmission of cooperation]
Without vertical transmission ($v=0$), if there is a horizontal transmission bias in favor of cooperation, namely
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
\end{result}

Therefore, in the absence of vertical transmission, selection plays no role in the evolution of cooperation. Hence, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~{\autoref{fig:results_c}).

@@

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
\tilde{p}(1-\tilde{p})\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.

Therefore, for cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$, that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_global_condition}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1 \Leftrightarrow
  c < \frac{bT_A + (T_A- T_B)}{1-T_B}.
\end{equation}
\\

If $\alpha<1$, we want to determine a condition for $\tilde{p}'>\tilde{p}$. 
We divide \autoref{eq:nextgen_parents_vertical_only} by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$ to get
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Rearranging, we get
\begin{equation} 
\begin{aligned} 
  1 - \tilde{p} < 
  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Diving by $(1-\tilde{p})$ and rearranging so that free terms are on the left and terms with $\tilde{p}$ are on the right, we have
\begin{equation} 
\begin{aligned} 
  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
   + (1-c) (1-\alpha) T_B
   + (1+b) (1-\alpha) T_A 
   - (1-\alpha) T_A].
\end{aligned}
\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

Following the same steps to solve $\tilde{p}'=\tilde{p}$, we find that there can be a third, polymorphic equilibrium 
\begin{equation} \label{eq:vert_hori_equilibrium}
  \tilde{p}^* = 
  \frac{c(1-T_B) - b \alpha T_A - (T_A-T_B)}{b (1-\alpha) (T_A-T_B)} .
\end{equation} 
Note that this is a legitimate equilibrium only if $0<\tilde{p}^*<1$.

Note that all parameters are positive.
So, applying \autoref{eq:vert_hori_global_condition}, for $\tilde{p}'>\tilde{p}$ we require that either 
\begin{align} 
\label{eq:vert_hori_global_condition1}
T_A > T_B \quad&\text{and}\quad \tilde{p}>\tilde{p}^*,  \quad \text{or} \\
\label{eq:vert_hori_global_condition2}
T_A <T_B \quad&\text{and}\quad \tilde{p}<\tilde{p}^* .
\end{align}

@@

\begin{result}[Vertical and horizontal transmission of cooperation]
Without oblique transmission ($v=1$), fixation, extinction, and coexistence of both phenotypes are possible.
\end{result}

We define the initial frequency as $\tilde{p}_0$ and
\begin{equation}\begin{aligned}
\gamma_1 = \frac{b \alpha T_A - (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}\end{equation}
Applying eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2} we can summarize the possible outcomes:
\begin{enumerate}
\item Fixation of cooperation,
	if $T_A>T_B$ and $c < \gamma_1$, or
	if $T_A>T_B$ and $\gamma_1 < c < \gamma_2$, and $\tilde{p}_0>\tilde{p}^*$, or
	if $T_A<T_B$ and $\gamma_2 < c$.
\item Global fixation of defection,
	if $T_A>T_B$ and $\gamma_2 < c$, or
	if $T_A<T_B$ and $c < \gamma_1$, or
	if $T_A<T_B$ and $\gamma_1 < c < \gamma_2$, and $\tilde{p}_0<\tilde{p}^*$.
\item Co-existence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and
	$\gamma_2 < c < \gamma_1$.
\end{enumerate}

Importantly, the condition for cooperation to increase from rarity ($\tilde{p}_0 \approx 0$) is
\begin{equation} \label{eq:unequal_transmission_from_rarity}
T_A>T_B \; \text{and} \; c < \gamma_1
\end{equation} 
Thus, cooperation must have a horizontal transmission bias, and the cost of cooperation cannot be too big.

In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, we get the following corollary that does have the form of Hamilton's rule.\\

@@

\begin{corollary}[No assortment of transmission and cooperation]
When $\alpha=0$, 
if there is a horizontal transmission bias for cooperation, $T_A>T_B$, the cost is not too high $c < (1+b) \frac{T_A-T_B}{1-T_B}$, and the initial frequency is high enough, $\tilde{p}_0 > \tilde{p}^*$, then cooperation will fix in the population,
If there is a horizontal transmission bias for defection, $T_A<T_B$, defection will take over the population.
\end{corollary}

Here, the third equilibrium is
\begin{equation} \label{eq:vert_hori_alpha0_equilibrium}
\tilde{p}^*(\alpha=0) = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B) },
\end{equation} 
and the cost boundaries are
\begin{equation}\begin{aligned}
\gamma_1(\alpha=0) = -\frac{(T_A - T_B)}{1-T_B}, \quad
\gamma_2(\alpha=0) = \frac{(1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}\end{equation}
If $T_A>T_B$ then $c>0>\gamma_1$ and cooperation cannot increase from rarity, but it can fix if $c<\frac{(1+b) (T_A - T_B)}{1-T_B}$; otherwise defection will fix.

If $T_A<T_B$ then % TODO check this... gamma2<0 means fixation of A???
\\
